Extraction of programs from proofs

### 2022-10-02 slides-fischbachau2022.tex
Notes almost finished

\title{Revisiting divisible, injective and flabby abelian groups from a
constructive point of view}
\author{Ingo Blechschmidt}
\date{September 21th, 2022}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1-lighter}\end{minipage}}
\begin{frame}{}
  \medskip
  \textbf{Thm.}
  For every number~$n \in \NN$, there is a prime larger than~$n$.

@@

  {\centering\emph{``Every theorem has a computable* witness.''} \\ \scriptsize
  * with monadic side effects\par}
\end{frame}}

@@

    * 1815 \ \ † 1852
  \end{minipage}
\end{frame}}

@@

  \emph{for extracting programs from constructive proofs}
\end{frame}

@@

  \hil{Heyting arithmetic} has as axioms (the universal closure of)
  \begin{align*}
    &&& \neg(0 = Sx) \\
    &&& S(x) = S(y) \Rightarrow x = y \\
    x + 0 &= x &&&
    x \cdot 0 &= 0 \\
    x + S(y) &= S(x+y) &&&
    x \cdot S(y) &= (x \cdot y) + x
  \end{align*}
  together with the \hil{induction scheme} (one axiom for each
  formula~$\varphi$)
  \[
    \varphi(0) \wedge \bigl(\forall x\?N\_ \varphi(x) \Rightarrow
    \varphi(S(x))\bigr) \quad\Longrightarrow\quad \forall x\?N\_ \varphi(x)
  \]
  and the rules of \hil{sequence calculus}.
\end{frame}

@@

\begin{frame}{Number realizability}
  \vspace*{-1em}
  \begin{changemargin}{-1.5em}{-0.5em}
  \begin{tabbing}
    $e \models (\forall f\?\NN^\NN\_ \varphi(n))$ \= \kill
    $e \realizes s = t$ \> iff $s = t$. \\
    $e \realizes \top$ \> iff true. \\
    $e \realizes \bot$ \> iff false. \\
    $e \realizes (\varphi \wedge \psi)$ \> iff~$\pi_1 \cdot e \downarrow$
    and~$\pi_2 \cdot e \downarrow$ and $\pi_1 \cdot e \realizes \varphi$
    and~$\pi_2 \cdot e \realizes \psi$. \\
    $e \realizes (\varphi \vee \psi)$ \> iff~$\pi_1 \cdot e \downarrow$
    and~$\pi_2 \cdot e \downarrow$ and \\ \> \qquad if~$\pi_1 \cdot e = 0$
    then~$\pi_2 \cdot e \realizes
    \varphi$, and \\ \> \qquad if~$\pi_1 \cdot e \neq 0$ then~$\pi_2 \cdot e \realizes \psi$. \\
    $e \realizes (\varphi \Rightarrow \psi)$ \> iff for every~$r \in \NN$
    such that~$r \realizes \varphi$, $e \cdot r \downarrow$ and~$e \cdot r \realizes \psi$. \\
    $e \realizes (\forall n\?N\_ \varphi(n))$ \> iff for every~$n_0
    \in \NN$, $e \cdot n_0 \downarrow$ and~$e \cdot n_0 \realizes \varphi(n_0)$. \\
    $e \realizes (\exists n\?N\_ \varphi(n))$ \> iff~$\pi_1 \cdot e
    \downarrow$ and~$\pi_2 \cdot e \downarrow$
    and~$\pi_2 \cdot e \realizes \varphi(\pi_1 \cdot e)$. \\
    $e \realizes (\forall f\?N^N\_ \varphi(f))$ \> iff for every~$f_0
    : \NN \to \NN$ and every~$r_0 \in \NN$ such that \\ \> \qquad $f_0$ is computed by the~$r_0$-th
    machine, \\ \> \qquad
    $e \cdot r_0 \downarrow$ and~$e \cdot r_0 \realizes \varphi(f_0)$. \\
    $e \realizes (\exists f\?N^N\_ \varphi(f))$ \> iff~$\pi_1 \cdot e \downarrow$
    and~$\pi_2 \cdot e \downarrow$ and
    the $(\pi_1 \cdot e)$-th machine \\ \> \qquad computes a function~$f_0 : \NN \to \NN$
    and $\pi_2 \cdot e \realizes \varphi(f_0)$.
  \end{tabbing}\end{changemargin}

  \mbox{\textbf{Thm.} If~$\text{HA} \vdash \varphi$, then there is a
  number~$e \in \NN$ such that~\only<1>{$e \realizes
  \varphi$}\only<2>{$\text{HA} \vdash (\underline{e} \realizes \varphi)$}.}
\end{frame}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{0.00cm}\hfill\mbox{\includegraphics[width=0.2\textwidth]{turing-machine}\;\;}\end{minipage}}
\begin{frame}{Exploring the realizability model}
  \medskip
  \medskip
  \medskip
  \begin{tabular}{@{\!\!\!\!\!\!}l@{\,}llp{1.8cm}}
    \toprule
    & statement & classical? & realizable? \\
    \midrule
    \normalnumber{1} & Every number is prime or not prime. & \ccmark{}
    (trivially) & \ccmark \\
    \normalnumber{2} & After every number there is a prime. & \ccmark & \ccmark \\
    \normalnumber{3} & Every map $\NN \to \NN$ has a zero or not. & \ccmark{} (trivially) & \cxmark \\
    \normalnumber{4} & Every map $\NN \to \NN$ is computable. & \cxmark &
    \qswitch{4}{5}{\ccmark}\only<1-4>{\,} \visible<5->{(trivially)} \\
    \normalnumber{5} & Every map $\RR \to \RR$ is continuous. & \cxmark &
    \qswitch{5}{6}{\ccmark} \\
    \normalnumber{6} & Markov's principle holds. & \ccmark{} (trivially) &
    \qswitch{6}{7}{\ccmark{} (if MP)} \\
    \normalnumber{7} & Countable choice holds. & \ccmark &
    \qswitch{7}{8}{\ccmark{} (always!)} \\
    \normalnumber{8} & Heyting arithmetic is categorical. & \cxmark &
    \qswitch{8}{9}{\ccmark{} (if MP)} \\
    \bottomrule
  \end{tabular}
  \medskip

  \only<1>{\ \\\ \\\ \\\ }
  \only<2>{\expl{1}{There is a machine which determines of any given
  number whether it is prime or not. \\\ }}
  \only<3>{\expl{2}{There is a machine which, given a number~$n$, computes a
  prime larger than~$n$. \\\ \\\ }}
  \only<4>{\expl{3}{There is a machine which, given a machine
  computing a map~$f : \NN \to \NN$, determines whether~$f$ has a
  zero or not. \\\ }}
  \only<5>{\expl{4}{There is a machine which, given a machine
  computing a map~$f : \NN \to \NN$, outputs a machine
  computing~$f$. \\\ }}
  \only<6>{\ \\\ \\\ \\\ }
  \only<7>{\expl{6}{There is a machine which, given a machine
  computing a map~$f : \NN \to \NN$ and given the promise that it is \notnot
  the case that~$f$ has a zero, determines a zero of~$f$.}}
  \only<8->{\expl{7}{\justifying There is a machine which, given a machine
  computing for every~$x \in \NN$ some~$y \in A$ together with a realizer
  of~$\varphi(x,y)$, outputs a machine computing a suitable choice
  function~$\NN \to A$.}}
\end{frame}}

\begin{frame}{Metatheory of Heyting arithmetic}
  \begin{enumerate}
    \item \hil{Unprovability results:}
    \medskip

@@

\begin{frame}{Range of machine models}
  \begin{enumerate}
    \item \hil{Turing machines}

@@

    \item \hil{Machines in the real world} -- philosophical

@@

      \textbf{Philosopher.} Thank you. \emph{[She inspects the stone.]} This is
      indeed the Philosopher's Stone. Many years ago you asked me to either
      acquire the Philosopher's Stone or find out how to produce arbitrary
      amounts of gold using it. Now it's my pleasure to present to you the
      Philosopher's Stone. \emph{[She returns the stone.]}
    \end{multicols}
  \end{changemargin}
\end{frame}

@@

\begin{frame}{A case study in double negation}
  \vspace*{-0.8em}
  \mbox{How to extract \good{constructive proofs} from
  \bad{classical proofs} as the following?}

@@

\begin{frame}{The double-negation embedding}
  \justifying
  \vspace*{-1em}
  \small
  \textbf{Def.} For formulas over a fixed first-order signature, the
  \hil{$\neg\neg$-translation} $\varphi \mapsto \varphi^{\neg\neg}$ is defined
  by the following clauses.
  \begin{align*}
    (\varphi_\text{atomic})^{\neg\neg} &\defeqv \hnegneg \varphi_\text{atomic} &
    (\varphi \Rightarrow \psi)^{\neg\neg} &\defeqv (\varphi^{\neg\neg} \Rightarrow \psi^{\neg\neg}) \\
    \bot^{\neg\neg} &\defeqv \hnegneg\bot &
    \top^{\neg\neg} &\defeqv \top \\
    (\varphi \vee \psi)^{\neg\neg} &\defeqv \hnegneg(\varphi^{\neg\neg} \vee \psi^{\neg\neg}) &
    (\varphi \wedge \psi)^{\neg\neg} &\defeqv (\varphi^{\neg\neg} \wedge \psi^{\neg\neg}) \\
    (\exists x\?X\_ \varphi)^{\neg\neg} &\defeqv \hnegneg{(\exists x\?X\_ \varphi^{\neg\neg})} &
    (\forall x\?X\_ \varphi)^{\neg\neg} &\defeqv (\forall x\?X\_ \varphi^{\neg\neg}) \\
  \end{align*}
  \vspace*{-3em}

>>>>

\title{Extraction of programs from proofs}
\author{Ingo Blechschmidt}
\date{September 21th, 2022}

\setbeameroption{show notes on second screen}
%\setbeamertemplate{note page}[plain, bg=white!90!black]
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1-lighter}\end{minipage}}
\begin{frame}{}
  \jnote{1-5}{
    From the displayed proof of Euclid's theorem, we can read off an algorithm
    for computing arbitrarily large primes. There is a deeper reason to that:
    The proof is \emph{constructive}, and from \emph{every} constructive proof
    we can extract a corresponding program. One way to formally state and prove
    this meta-statement is by \emph{realizability theory}, the subject of the
    first lecture.
  }

  \jnote{4-5}{
    In addition to the displayed applications of realizability theory,
    personally I'm intrigued by it for mostly the following reasons:
    (1)~Realizability elucidates the interplay between constructive and
    computable mathematics. (2)~Realizability is a useful guide for pursuing
    the question whether two given proofs are ``secretly the same'': Do they
    have the same computational content? (3)~Realizability theory provides us
    with a host of tantalizing anti-classical models of constructive
    mathematics.
  }

  \jnote{5-5}{
    In the second lecture, we will turn to extracting programs from
    \emph{classical} proofs. We will do so by transforming classical proofs
    into constructive ones and then applying the tools of the first lecture.
    Amazingly, the displayed classical proof and others like it do have
    constructive content---even though it is constructively and computably
    impossible to determine minimal values of infinite sequences.
  }

  \jnote{6-}{
    Finally, in the third lecture we will learn how to extract constructive
    content from certain kinds of \emph{invalid} proofs---those which use the
    preposterous assumption that every set is countable.

    The methods presented in the second and in the third lecture are deeply
    related to the \emph{dynamical approach to algebra} reported on in Stefan
    Neuwirth's course. Coherent (and geometric) logic as in Marc Bezem's course
    also plays an important role in this toolbox. It is greatly informed by a
    categorical analysis as provided in Steve Awodey's course, particularly so
    for the third lecture. The first and the second lecture overlap with
    Chuangjie Xu's course, particularly in employing variants of the
    double-negation translation.

    This course is set in an informal constructive metatheory. Formalization
    would both be possible in type theory as in Fredrik Nordvall-Forsberg's
    course or in constructive set theory as in Hajime Ishihare's course.
    All invocations of the transfinite such as by the law of excluded middle or
    by the axiom of choice will be explicitly pointed out. For primers to
    constructive mathematics, enjoy
    \fixedhref{https://video.ias.edu/members/1213/0318-AndrejBauer}{Andrej
    Bauer's 2013 IAS talk}, its \fixedhref{xxx}{written version} or
    \fixedhref{xxx}{these course notes}.
  }

  \medskip
  \textbf{Thm.}
  For every number~$n \in \NN$, there is a prime larger than~$n$.

@@

  {\centering\emph{``\textnormal{Every} theorem has a computable* witness.''} \\ \scriptsize
  * with monadic side effects\par}
\end{frame}}

@@

    * 1815 \ \ † 1852
  \end{minipage}

  \jnote{1}{
    This year we will celebrate the 207th birthday of Ada Lovelace, pioneer in
    computing.

    It is astonishing what she started and which long way we have come!

    Perhaps you would enjoy the graphic novel \emph{The Thrilling Adventures of
    Lovelace and Babbage} in her honor.
  }
\end{frame}}

@@

  \emph{for extracting programs from constructive proofs}

  \jnote{1}{
    For a written primer to realizability theory, see
    \fixedhref{http://math.andrej.com/asset/data/c2c.pdf}{Andrej Bauer's course
    notes} and the notes by
    \fixedhref{https://www2.mathematik.tu-darmstadt.de/~streicher/REAL/REAL.pdf}{Thomas
    Streicher}.
  }
\end{frame}

@@

  \hil{Heyting arithmetic} has as axioms (the universal closure of)
  \begin{align*}
    &&& \neg(0 = Sx) \\
    &&& S(x) = S(y) \Rightarrow x = y \\
    x + 0 &= x &&&
    x \cdot 0 &= 0 \\
    x + S(y) &= S(x+y) &&&
    x \cdot S(y) &= (x \cdot y) + x
  \end{align*}
  together with the \hil{induction scheme} (one axiom for each
  formula~$\varphi$)
  \[
    \varphi(0) \wedge \bigl(\forall x\?N\_ \varphi(x) \Rightarrow
    \varphi(S(x))\bigr) \quad\Longrightarrow\quad \forall x\?N\_ \varphi(x)
  \]
  and the rules of \hil{sequence calculus}.

  \jnote{1}{
    Heyting arithmetic is a convenient base theory for a constructive analysis
    of arithmetic. It has exactly the same axioms as Peano arithmetic, only
    that HA is set in intuitionistic logic while PA adds the law of excluded
    middle.

    As is common, we define negation~``$\neg\varphi$'' as a shorthand for the
    implication~``$\varphi \Rightarrow \bot$''.

    HA is often expanded to HA\textsuperscript{$\omega$}, \emph{higher-order
    Heyting arithmetic}, which includes sorts, term constructors and
    appropriate rules for function types such as~$N^N$ and $N^{(N^N)}$.

    In its original form, realizability theory is only concernced with
    extracting computational witnesses from HA-proofs; however it is fruitfully
    extended to all of~HA\textsuperscript{$\omega$}, and we will also glimpse
    into this higher-order extension. (NB: HA\textsuperscript{$\omega$} is
    conservative over~HA, and one way to show this is by realizability.)
  }
\end{frame}

@@

\begin{frame}{Number realizability}
  \jnote{1}{
    A statement~$\varphi$ is \emph{realizable}, written~``$\realizes\varphi$'',
    iff it has a \emph{realizer}, a number~$e \in \NN$ such that~$e \realizes
    \varphi$. The recursive rules governing which numbers~$e$ are deemed to be
    a realizer of~$\varphi$ make use of \emph{Kleene's original partial combinatory
    algebra}, the natural numbers equipped with the following partial binary
    operation~$({\cdot})$: $e \cdot n$ is the result of applying the
    input~$n$ to the~$e$-th Turing machine (in some effective enumeration of all
    Turing machines). We write~``$e \cdot n \downarrow$'' to signify that this
    computation terminates.

    Instead of Turing machines, we can also study other deterministic models of
    computation by using different partial combinatory algebras, or even
    nondeterministic and stateful models by a recent
    \fixedhref{https://rosstate.org/publications/effectful/effectful-mfps19.pdf}{tantalizing
    generalization} due to Liron Cohen and her coauthors Sofia Abreu Faro and
    Ross Tate.

    Realizability theory provides one way of formalizing the informal
    Brouwer--Heyting--Kolmogorov interpretation of constructive mathematics.
    For instance, that interpretation states that a witness of an
    implication~$\varphi \Rightarrow \psi$ is a ``method'' for transforming
    witnesses for~$\varphi$ into witnesses for~$\psi$. Realizability spells out
    what ``method'' should mean: Turing machine.
  }

  \jnote{2}{
    The clauses for disjunction and existential quantification require pairing
    and unpairing. Given two numbers~$a$ and~$b$, there is a Turing
    machine~$p_{a,b}$ which outputs~$a$ or~$b$ depending on whether its input
    is zero or not zero. By~$\pi_1$ and~$\pi_2$, we mean (indices of) Turing
    machines which, when called on input (an index of)~$p_{a,b}$, extract~$a$
    repectively~$b$ by simulating its input on the input~$0$ or~$1$.

    The soundness theorem is proven by an instructive induction on the
    structure of~HA-proofs, verifying that if HA proves a sequent~$\varphi
    \seq{\vec x} \psi$, then there is a realizer for~$\forall x_1\_ \ldots
    \forall x_n\_ (\varphi \Rightarrow \psi)$. The core idea of the proof is to
    verify that every axiom and every rule of~HA is realized. In this way,
    computational content of every axiom and every rule is explicated.
  }

  \jnote{3}{
    In the form ``$(\text{HA} \vdash \varphi) \Rightarrow
    (\realizes\varphi)$'', the soundness theorem can be stated and proved in
    most contexts in which the natural numbers exist as a complete entity, such
    as constructive or classical set or type theories.

    But in a sense, this is misleading: The mapping from proofs to realizers is
    a computationally simple syntactical transformation. As such, already PRA
    can prove the soundness theorem if we formulate it as~``$(\text{HA} \vdash
    \varphi) \Rightarrow (\exists e\_ \text{HA} \vdash (e\realizes\varphi))$''.

    NB: Some formulations of realizability state the clauses for disjunction
    and existential quantification in a slighter simpler way, directly using
    pairing and unpairing functions on the naturals. The price for this
    simplification is that then the soundness theorem has to be formulated
    as``$(\text{HA} \vdash \varphi) \Rightarrow (\exists e\_ \text{HA} \vdash
    (e \realizes(\top \Rightarrow \varphi)))$''.
  }

  \vspace*{-1em}
  \begin{changemargin}{-1.5em}{-0.5em}
  \begin{tabbing}
    $e \models (\forall f\?\NN^\NN\_ \varphi(n))$ \= \kill
    $e \realizes s = t$ \> iff $s = t$. \\
    $e \realizes \top$ \> iff true. \\
    $e \realizes \bot$ \> iff false. \\
    $e \realizes (\varphi \wedge \psi)$ \> iff~$\pi_1 \cdot e \downarrow$
    and~$\pi_2 \cdot e \downarrow$ and $\pi_1 \cdot e \realizes \varphi$
    and~$\pi_2 \cdot e \realizes \psi$. \\
    $e \realizes (\varphi \vee \psi)$ \> iff~$\pi_1 \cdot e \downarrow$
    and~$\pi_2 \cdot e \downarrow$ and \\ \> \qquad if~$\pi_1 \cdot e = 0$
    then~$\pi_2 \cdot e \realizes
    \varphi$, and \\ \> \qquad if~$\pi_1 \cdot e \neq 0$ then~$\pi_2 \cdot e \realizes \psi$. \\
    $e \realizes (\varphi \Rightarrow \psi)$ \> iff for every~$r \in \NN$
    such that~$r \realizes \varphi$, $e \cdot r \downarrow$ and~$e \cdot r \realizes \psi$. \\
    $e \realizes (\forall n\?N\_ \varphi(n))$ \> iff for every~$n_0
    \in \NN$, $e \cdot n_0 \downarrow$ and~$e \cdot n_0 \realizes \varphi(n_0)$. \\
    $e \realizes (\exists n\?N\_ \varphi(n))$ \> iff~$\pi_1 \cdot e
    \downarrow$ and~$\pi_2 \cdot e \downarrow$
    and~$\pi_2 \cdot e \realizes \varphi(\pi_1 \cdot e)$. \\
    $e \realizes (\forall f\?N^N\_ \varphi(f))$ \> iff for every~$f_0
    : \NN \to \NN$ and every~$r_0 \in \NN$ such that \\ \> \qquad $f_0$ is computed by the~$r_0$-th
    machine, \\ \> \qquad
    $e \cdot r_0 \downarrow$ and~$e \cdot r_0 \realizes \varphi(f_0)$. \\
    $e \realizes (\exists f\?N^N\_ \varphi(f))$ \> iff~$\pi_1 \cdot e \downarrow$
    and~$\pi_2 \cdot e \downarrow$ and
    the $(\pi_1 \cdot e)$-th machine \\ \> \qquad computes a function~$f_0 : \NN \to \NN$
    and $\pi_2 \cdot e \realizes \varphi(f_0)$.
  \end{tabbing}\end{changemargin}

  \mbox{\textbf{Thm.} If~$\text{HA} \vdash \varphi$, then there is a
  number~$e \in \NN$ such that~\only<1-2>{$e \realizes
  \varphi$}\only<3>{$\text{HA} \vdash (\underline{e} \realizes \varphi)$}.}
\end{frame}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{0.00cm}\hfill\mbox{\includegraphics[width=0.2\textwidth]{turing-machine}\;\;}\end{minipage}}
\begin{frame}{Exploring the realizability model}
  \jnote{5}{
    The statement that every function~$\NN \to \NN$ is computable by a Turing
    machine (or equivalently, by a lambda term) is known as the \emph{formal
    Church--Turing thesis}. It is an example of a statement which is realizable
    but not provable in~HA\textsuperscript{$\omega$}.

    Many of the curious properties of the realizability model follow from the
    formal Church--Turing thesis. In fact, a slight generalization called the
    \emph{extended Church thesis} suffices to completely characterize the
    (provably) realizable statements:
    \[ \text{HA+ECT} \vdash \varphi \quad\text{iff}\quad
      \text{HA} \vdash (\realizes \varphi). \]

    In the realizability model built using lambda terms instead of Turing
    machines, the formal Church--Turing thesis fails. This is because of
    changed calling conventions: In the model built using lambda terms, a
    realizer for a statement of the form~``$\forall f\?N^N\_ \varphi(f)$'' is a
    lambda term~$e$ such that for every lambda term~$r$ computing a
    function~$f : \NN \to \NN$, the term $er$ is a realizer for~$\varphi(f)$.
    However, the term~$e$ cannot inspect the form (``source code'') of its
    argument.
  }

  \jnote{6}{
    Statement~5 is not a statement in the language of~HA or
    of~HA\textsuperscript{$\omega$}, but in an extension in which we can also
    support quotients (to make sense of the construction of the reals using
    equivalence classes of Cauchy sequences) or powersets (to support the
    construction using Dedekind cuts). A proper interpretation is possible in
    the category of assemblies or in the ef{}fective topos.

    An exposition of this continuity phenomenon is provided in
    \fixedhref{https://arxiv.org/pdf/2204.00948.pdf}{this survey paper}
    (Example~6 there).
  }

  \jnote{7}{
    By \emph{Markov's principle}, we mean the statement
    \[ \forall f\?N^N\_ (\neg\neg(\exists n\?N\_ f(n) = 0)) \Rightarrow
    (\exists n\?N\_ f(n) = 0). \]
    By the clauses for implication and negation, a number~$e$ is a realizer for
    a negated statement~$\neg\psi$ iff there is no realizer for~$\psi$:
    \begin{align*}
      e \realizes \neg\psi &\quad\text{iff}\quad
      \text{for every~$r \in \NN$ such that~$r \realizes \psi$, $e \cdot r
      \downarrow$ and~$e \cdot r \realizes \bot$} \\
      &\quad\text{iff}\quad \text{for every~$r \in \NN$ such that~$r \realizes \psi$, falsum holds} \\
      &\quad\text{iff}\quad \text{there is no number~$r \in \NN$ such that~$r \realizes \psi$} \\
      &\quad\text{iff}\quad \text{$\psi$ is not realized}
    \end{align*}
    In particular, if there exists a realizer for a negated statement at all,
    every number whatsoever is a realizer. As a consequence, \emph{realizers for
    negated statements are never informative;} and a number~$e$ is a realizer
    for~$\neg\neg\varphi$ iff~$\varphi$ is \emph{not~not} realizable. Hence a
    realizer for~$\neg\neg\varphi$ encodes the mere promise that somewhere,
    there is a realizer for~$\varphi$, without giving any indication how to
    find it.
  }

  \jnote{10}{
    As the examples illustrate, realizability and truth do not at all coincide:
    There are many statements which are realizable but not true from the point of
    view of classical mathematics (such as the formal Church--Turing thesis)
    and vice versa (such as the statement that every function~$\NN \to \NN$ has
    a zero or not).

    Within the realizability model, the situation is radically different.
    For every statement~$\varphi$, the statement~``$\varphi \Leftrightarrow
    (\realizes \varphi)$'' is realizable.

    The multiverse of models of constructive mathematics can be explored from
    the point of view of any base model, and from the point of view of the
    realizability model it looks quite different that from the point of view of
    classical mathematics.
  }

  \medskip
  \medskip
  \begin{tabular}{@{\!\!\!\!\!\!}l@{\,}llp{1.8cm}}
    \toprule
    & statement & classical? & realizable? \\
    \midrule
    \normalnumber{1} & Every number is prime or not prime. & \ccmark{}
    (trivially) & \ccmark \\
    \normalnumber{2} & After every number there is a prime. & \ccmark & \ccmark \\
    \normalnumber{3} & Every map $\NN \to \NN$ has a zero or not. & \ccmark{} (trivially) & \cxmark \\
    \normalnumber{4} & Every map $\NN \to \NN$ is computable. & \cxmark &
    \qswitch{4}{5}{\ccmark}\only<1-4>{\,} \visible<5->{(trivially)} \\
    \normalnumber{5} & Every map $\RR \to \RR$ is continuous. & \cxmark &
    \qswitch{5}{6}{\ccmark{} (if MP)} \\
    \normalnumber{6} & Markov's principle holds. & \ccmark{} (trivially) &
    \qswitch{6}{7}{\ccmark{} (if MP)} \\
    \normalnumber{7} & Countable choice holds. & \ccmark &
    \qswitch{7}{8}{\ccmark{} (always!)} \\
    \normalnumber{8} & Heyting arithmetic is categorical. & \cxmark &
    \qswitch{8}{9}{\ccmark{} (if MP)} \\
    \normalnumber{9} & A statement holds iff it is realized. & \cxmark &
    \qswitch{9}{10}{\ccmark} \\
    \bottomrule
  \end{tabular}
  \medskip

  \only<1>{\color{white}There is a machine which determines of any given
  number whether it is prime or not. \\\ \\\ }
  \only<2>{\expl{1}{There is a machine which determines of any given
  number whether it is prime or not. \\\ \\\ }}
  \only<3>{\expl{2}{There is a machine which, given a number~$n$, computes a
  prime larger than~$n$. \\\ \\\ }}
  \only<4>{\expl{3}{There is a machine which, given a machine
  computing a map~$f : \NN \to \NN$, determines whether~$f$ has a
  zero or not. \\\ \\\ }}
  \only<5>{\expl{4}{There is a machine which, given a machine
  computing a map~$f : \NN \to \NN$, outputs a machine
  computing~$f$. \\\ \\\ }}
  \only<6>{\ \\\ \\\ \\\ }
  \only<7>{\expl{6}{There is a machine which, given a machine
  computing a map~$f : \NN \to \NN$ and given the promise that it is \notnot
  the case that~$f$ has a zero, determines a zero of~$f$.}}
  \only<8>{\expl{7}{\justifying There is a machine which, given a machine
  computing for every~$x \in \NN$ some~$y \in A$ together with a realizer
  of~$\varphi(x,y)$, outputs a machine computing a suitable choice
  function~$\NN \to A$.}}
  \only<9>{\ \\\ \\\ \\\ }
\end{frame}}

\begin{frame}{Metatheory of Heyting arithmetic}
  \jnote{1}{
    Variants of the realizability model can be used to establish several
    metatheoretic properties of Heyting arithmetic. For the second and third
    properties, the keyword is ``realizability with proof''; for the fourth,
    using the variant of realizability built using System~T terms instead of
    Turing machines.
  }

  \begin{enumerate}
    \item \hil{Unprovability results:}
    \medskip

@@

\begin{frame}{Range of machine models}
  \jnote{1-}{
    Infinite time Turing machines were introduced
    \fixedhref{https://arxiv.org/abs/math/9808093}{by Joel David Hamkins and
    Andy Lewis}. Unlike ordinary Turing machines, they can carry out ``more
    than infinitely many computational steps''. Where an ordinary Turing
    machine fails to terminate, an infinite time Turing machine is put on
    day~$\omega$ into a special limit state and can then meaningfully continue.

    All functions in Gödel's System~T are unconditionally total. Hence
    unbounded search cannot be implemented in System~T.

    The idea to apply realizability to machines in the real world is
    \fixedhref{http://math.andrej.com/2014/03/04/intuitionistic-mathematics-and-realizability-in-the-physical-world/}{due to Andrej Bauer}.
  }

  \begin{enumerate}
    \item \hil{Turing machines}

@@

    \item \hil{Machines in the real world} -- \emph{philosophical}

@@

      \textbf{Philosopher.} Thank you. \emph{[She inspects the stone.]} This is
      indeed the Philosopher's Stone. Many years ago you asked me to either
      acquire the Philosopher's Stone or find out how to produce arbitrary
      amounts of gold using it. Now it's my pleasure to present to you the
      Philosopher's Stone. \emph{[She returns the stone.]}
    \end{multicols}
  \end{changemargin}

  \jnote{1}{
    Adapted from
    \fixedhref{http://blog.ezyang.com/2013/04/a-classical-logic-fairy-tale/}{Edward
    Yang's blog}.
  }
\end{frame}

@@

\begin{frame}{A case study in double negation}
  \jnote{2-4}{
    The two displayed proofs seem to be devoid of computational content:
    Computationally, it is neither possible to determine the minimal element of
    an infinite sequence nor to determine whether a binary sequence contains
    infinitely many zeros or infinitely many ones. These claims have no
    realizer.

    However, contrary to expectations, the two proofs do contain an obscured
    constructive core, and this core can be uncovered by the two logical
    metatheorems presented in this lecture, the double-negation embedding and
    \emph{Friedman's trick}. Details are in Thierry Coquand's
    \fixedhref{https://www.lama.univ-savoie.fr/pagesmembres/ruyer/divers/lecturearticles/coquand/cor3.ps}{Computational
    Content of Classical Logic}.
  }

  \jnote{3-4}{
    To set the stage, we analyze the lemma used in the first proof. The
    statement ``every inhabited set of natural numbers contains a minimal
    element'' is not realizable and hence does not directly have
    computational content. In fact, this statement is equivalent to the law of
    excluded middle. Correspondingly, it seems that proofs using it cannot ever
    be constructivized. However, as we will see, much more important for
    extraction of constructive content is the form of the asserted claims than
    the form of the used auxiliary lemmas.
  }

  \jnote{5-}{
    A set~$X$ of natural numbers is \emph{detachable} if and only if for every
    number~$n \in \NN$, either~$n \in X$ or~$n \not\in X$. In classical
    mathematics, every set of natural numbers is detachable. Constructively,
    detachability is a nontrivial and computationally meaningful condition. A
    realizer for detachability of a set~$X$ is a machine which reads a
    number~$n$ as input and determines whether~$n$ belongs to~$X$ or not
    (together with corresponding realizers).
  }

  \jnote{7-}{
    Instead of strengthening the assumption, we can also weaken the conclusion.
    Statement~3 merely asserts that it is impossible that no minimum exists.
    The \emph{double-negation embedding} reviewed on the next slide explains
    why it can be expected a priori that this weakening allows us to give a
    constructive proof.
  }

  \vspace*{-0.8em}
  \mbox{How to extract \good{constructive proofs} from
  \bad{classical proofs} as the following?}

@@

\begin{frame}{The double-negation embedding}
  \jnote{1-3}{
    The starting point of the double-negation embedding is that, while the law
    of excluded middle is not available in intuitionistic logic, the double
    negation of every instance is: For every formula~$\varphi$, the formula
    \[ \neg\neg(\varphi \vee \neg\varphi) \]
    is an intuitionistic tautology:

    \begin{quote}
      In order to show~$\neg\neg(\varphi \vee \neg\varphi)$,
      assume~$\neg(\varphi \vee \neg\varphi)$ and deduce~$\bot$.

      As a preparatory step, we verify~$\neg\varphi$: If~$\varphi$, then in
      particular~$\varphi \vee \neg\varphi$, hence~$\bot$ by assumption.

      Having established~$\neg\varphi$, we notice that also~$\varphi \vee
      \neg\varphi$, hence~$\bot$ by assumption.
    \end{quote}

%    This observation unlocks the following technique to transcribe a classical
%    proof of a statement to an intuitionistic (and even minimal logic) proof of
%    its translation. Most steps in the classical proof will also be acceptable
%    in intuitionistic logic. We simply mimic these. When the classical proof
%    appeals to~\textsc{lem}, we instead use the appropriate doubly negated
%    instance. We
%
%    The difference between intuitionistic and classical logic is concentrated
%    in their treatment of disjunction and existential quantification.
%    Classically, to verify a disjunction~$\varphi \vee \psi$, it suffices to
%    establish~$\neg(\neg\varphi \wedge \neg\psi)$--the intuitionistically
%    weaker claim that it is not the case that neither~$\varphi$ nor~$\psi$
%    hold; similarly, to verify an existential statement~$\exists x\?X\_
%    \varphi(x)$, it suffices to establish~$\neg\forall x\?X\_
%    \neg\varphi(x)$---the claim that nonexistence is impossible.
  }

  \jnote{3}{
    The three parts of the theorem are each proven by induction, on the
    structure of~$\varphi$ for the first two parts and on the structure of
    derivations for the third part. Carrying this out is an instructive
    exercise!
  }

  \jnote{4}{
    The corollary follows from the theorem because HA proves the
    double-negation translation of every axiom of PA. For instance, the
    translation of the axiom
    \[ \forall x\?N\_ x + 0 = x \]
    is
    \[ \forall x\?N\_ \neg\neg(x + 0 = x), \]
    and this weaker statement is provable thanks to the intuitionistic (even
    minimal) tautology~$\varphi \Rightarrow \neg\neg\varphi$.

    Similarly, the translation of an instance of the induction scheme
    \[
      \varphi(0) \wedge \bigl(\forall x\?N\_ \varphi(x) \Rightarrow
      \varphi(S(x))\bigr) \quad\Longrightarrow\quad \forall x\?N\_ \varphi(x)
    \]
    is again an instance of the induction scheme:
    \[
      \varphi(0)^{\neg\neg} \wedge \bigl(\forall x\?N\_ \varphi(x)^{\neg\neg} \Rightarrow
      \varphi(S(x))^{\neg\neg}\bigr) \quad\Longrightarrow\quad \forall x\?N\_
      \varphi(x)^{\neg\neg}
    \]
  }

  % xxx importance of remark
